Higher-Order unification for free!

### 2024-04-12 paper/main.tex
solve todo code
remove todo
abstract with spellchecker

\begin{abstract}
  Specifying and implementing a proof system from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are for free when ML binders
  represent object logic ones; 2) proof construction, and even proof search, is
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
  Constructions~\cite{felty93lics}.

  The object logic we are interested in is % an extension of the latter,
  Coq's~\cite{Coq-refman} Calculus of Inductive Constructions (CIC), for which 
  we want to implement a higher-order unification-based 
  proof search procedure using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory comprises
  $\eta\beta$ equivalence and comes equipped with a
  higher-order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  Elpi comes with an encoding of CIC that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
  but restricts \Ue to roughly first-order unification problems only.
  We call this basic encoding \Fo.

  In this paper we propose a better-behaved encoding \Ho,
  and show how to map unification problems in \Fo{}
  to related problems in \Ho.
    As a result we obtain \Uo{}, a higher-order unification procedure for \Fo{}
    that honours $\eta\beta$-equivalence (for CIC functions), solves
  problems in the pattern fragment and allows
  for the use of heuristics to deal with problems outside the
  pattern fragment. Moreover, since \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the ML, such as clause indexing.

@@

\input{code/comp_base}
\todo{use the real code}
\begin{elpicode}
type compile fm -> tm -> mmap -> mmap -> links -> links ->
  subst -> subst -> o.
compile Fo0 Ho M0 M1 L0 M1 S0 S1 :-
  beta-normal Fo0 Fo, comp Fo Ho M0 M1 L0 M1 S0 S1.
\end{elpicode}

@@

\input{code/unif_fo}
\todo{Fix real code calling compile (not comp), rename decompm to decomp}

>>>>

\begin{abstract}
  Specifying and implementing a proof system from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are for free when ML binders
  represent object logic ones; 2) proof construction, and even proof search, are
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
  Constructions~\cite{felty93lics}.

  The object logic we are interested in is
  Coq's Calculus of Inductive Constructions (CIC)\cite{Coq-refman}. We aim to
  develop a higher-order unification-based 
  proof search procedure for it using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi's equational theory includes
  $\eta\beta$ equivalence and features a
  higher-order unification procedure \Ue restricted to the pattern
  fragment~\cite{miller92jsc}.
  Elpi offers an encoding of CIC suitable
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}
  but restricts \Ue to roughly first-order unification problems only.
  We refer to this basic encoding as \Fo.

  In this paper we propose a more well-behaved encoding called \Ho,
  and show how to translate unification problems from \Fo{}
  to corresponding ones in \Ho.
  Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
    that honours $\eta\beta$-equivalence (for CIC functions), addresses
  problems within the pattern fragment, and allows
  for the use of heuristics to deal with problems outside the
  pattern fragment. Moreover, as \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the ML, such as clause indexing.

@@

\input{code/comp_base}
\input{code/compile.tex}

@@

\input{code/unif_fo}
